Manuscript Figures

### 2021-06-10 src/manuscript_figures.Rmd
update pancancer type prediction figure panels

```{r pancancer_prediction, fig.height=9, fig.width=12}
# Find out if cancer types are predictable using latent factors

res <- sapply(simplify = F, names(LFs), function(tool) {
  message(date(), " => ", tool)
  M <- LFs[[tool]]$pancancer.cnv_gex_meth_mut
  ct <- gsub("-", "_", surv[match(rownames(M), bcr_patient_barcode)]$project)
  res <- predict_labels_glmnet_multiclass(M, ct, partition = 0.6, nodes = 20)
})

@@

# performance visualisation
m <- as.matrix(table(res$maui_factors$pred, res$maui_factors$obs))
p2 <- pheatmap(apply(m, 1, function(x) round(x/sum(x) * 100, 1)), cluster_rows = F, cluster_cols = F, 
         display_numbers = T, number_format = "%.1f", silent = T)

@@

@@

# data type contribution to top LF per factor
p4 <- plot_datatype_contribution_by_factor(LFs, FWs, 'maui_factors', 
                                           'pancancer', 'cnv_gex_meth_mut', factors = ct, 
                                           top_lfs = 1, top_features = 100) 
p4 <- p4 + labs(y = 'Relative Contribution\nto top latent factor') + 
  theme(legend.position = 'right', legend.background = element_blank())

Fig <- plot_grid(plot_grid(p1, p2$gtable, nrow = 1, labels = c('A', 'B')), 
                  plot_grid(p3, p4, nrow = 1, labels = c('C', 'D')), 
                  nrow = 2,
                  rel_heights = c(2,3))

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.pancancer_cancer_type_prediction.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(file.path(FigureFolder, fig.name),
              plot = Fig, units = 'in', width = 12, height = 9)
Legends[[fig_count]] <- "Predictive power of latent factors in distinguishing cancer types. A-B) comparing maui with PCA/MOFA/MCIA. AUC: Area under the ROC curve computed on the test dataset (30% of held-out data) trained on 70% of the dataset using a Random Forest model. Each point represents one cancer type (pairwise one-sided Wilcoxon test on AUCs, alternative = maui AUC is not greater than others) C) Adjusted Mutual Information (AMI) value computed by comparing unsupervised clustering labels (k-means algorithm, k set to 21) with the cancer types. D) Top maui LFs detected per cancer type. Size/color/opacity of the points represent the average neural activation value of the LF for the corresponding cancer type samples. E) Percentage of top feature types contributing to top predictive maui latent factor per cancer type."

@@

multi_omics_comb <- 'cnv_gex_meth_mut'
experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
cl <- parallel::makeCluster(10)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
cluster.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
  message("Clustering top ",tool)
  stats <- do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, 
                                            intersect(names(LFs[[tool]]), experiments), 
                    function(experiment) {
    message(experiment)
    source(settings$utility_script)
    library(data.table)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      #M <- remove_redundant_variables(M, perc = 99) 
      M <- scale(M)
    }

    # try topN from 5 to 25 (each tool may capture different amount of information 
    # by varying number of factors)
    stats <- do.call(rbind, lapply(seq(5, min(20, ncol(M)), 5), function(n) {
      message(n)
      M.sub <- t(subset_features_by_variance(t(M), topN = n))
      val <- c('internal') 
      if(ncol(M.sub) > 2) {val <- c('internal', 'stability')}
      res <- clValid::clValid(obj = M.sub,
                            nClust = 3:8,
                            clMethods = 'kmeans',
                            validation = val,
                            maxitems = 10000,
                            nstart = 1000, iter.max = 30)
      stats <- data.table::as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')]
      colnames(stats) <- c('metric', 'k', 'clustering_method', 'score')
      stats$top_factors <- n
      return(stats)
    }))
    stats$cancer_type <- strsplit(experiment, "\\.")[[1]][1]
    stats$datatypes <- strsplit(experiment, "\\.")[[1]][2]
    return(stats)
  }))
  stats$method <- tool
  return(stats)
}))
parallel::stopCluster(cl)

@@

p1 <- ggboxplot(dt1, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = FALSE, #todo change to true when mofa is finished
                     comparisons = comps) + 
  labs(title = 'Silhouette score', x = '') + 
  theme(legend.position = 'none')

dt2 <- cluster.stats[metric == 'APN'][datatypes == multi_omics_comb,
                                             .SD[which.min(score)],
                                             by = c('cancer_type', 'method', 'k')]
dt2$method <- toupper(gsub("_factors", "", dt2$method))
comps <- lapply(setdiff(unique(dt2$method), 'MAUI'), function(x) c('MAUI', x))
p2 <- ggboxplot(dt2, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = FALSE,  #todo change to true when mofa is finished
                     comparisons = comps) + 
  labs(title = 'Cluster Stability Score (APN)', x = '') + 
  theme(legend.position = 'none')

>>>>

```{r pancancer_prediction, fig.height=9, fig.width=12}
# Find out if cancer types are predictable using latent factors 
res <- sapply(simplify = F, names(LFs), function(tool) {
  message(date(), " => ", tool)
  M <- LFs[[tool]]$pancancer.cnv_gex_meth_mut
  ct <- gsub("-", "_", surv[match(rownames(M), bcr_patient_barcode)]$project)
  res <- predict_labels_glmnet_multiclass(M, ct, partition = 0.6, nodes = 20)
})

@@

# performance visualisation
#m <- as.matrix(table(res$maui_factors$pred, res$maui_factors$obs))
#p2 <- pheatmap(apply(m, 1, function(x) round(x/sum(x) * 100, 1)), cluster_rows = F, cluster_cols = F, 
#         display_numbers = T, number_format = "%.1f", silent = T)

@@

p2 <- plot_tsne(M, ct, show.labels = T) + 
  theme(legend.position = 'none')

@@

# data type contribution to top LF per factor
p4 <- plot_datatype_contribution_by_factor(LFs, FWs, 'maui_factors', 
                                           'pancancer', 'cnv_gex_meth_mut', factors = ct, 
                                           top_lfs = 1, top_features = 100) 
p4 <- p4 + labs(y = 'Relative Contribution\nto top latent factor') + 
  theme(legend.position = 'top', legend.background = element_blank(), 
        legend.title = element_blank())

Fig <- plot_grid(plot_grid(p1, p2, nrow = 1, labels = c('A', 'B'),
                           rel_widths = c(1,2)), 
                  plot_grid(p4, p3, nrow = 1, labels = c('C', 'D'), 
                            rel_widths = c(1,2)), 
                  nrow = 2,
                  rel_heights = c(2,3))

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.pancancer_cancer_type_prediction.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(file.path(FigureFolder, fig.name),
              plot = Fig, units = 'in', width = 12, height = 12)
Legends[[fig_count]] <- "Predictive power of latent factors in distinguishing cancer types. A-B) comparing maui with PCA/MOFA/MCIA. AUC: Area under the ROC curve computed on the test dataset (30% of held-out data) trained on 70% of the dataset using a Random Forest model. Each point represents one cancer type (pairwise one-sided Wilcoxon test on AUCs, alternative = maui AUC is not greater than others) C) Adjusted Mutual Information (AMI) value computed by comparing unsupervised clustering labels (k-means algorithm, k set to 21) with the cancer types. D) Top maui LFs detected per cancer type. Size/color/opacity of the points represent the average neural activation value of the LF for the corresponding cancer type samples. E) Percentage of top feature types contributing to top predictive maui latent factor per cancer type."

@@

multi_omics_comb <- 'cnv_gex_meth_mut'
experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
cluster.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
  message("Clustering top ",tool)
  stats <- do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, 
                                            intersect(names(LFs[[tool]]), experiments), 
                    function(experiment) {
    message(experiment)
    source(settings$utility_script)
    library(data.table)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      # to avoid highly correlated maui LFs to dominate clusters
      M <- scale(t(remove_redundant_variables(t(M))))
    } 
    # try topN from 5 to 25 (each tool may capture different amount of information 
    # by varying number of factors)
    stats <- do.call(rbind, lapply(seq(5, min(20, ncol(M)), 5), function(n) {
      message(n)
      M.sub <- t(subset_features_by_variance(t(M), topN = n))
      val <- c('internal') 
      if(ncol(M.sub) > 2) {val <- c('internal', 'stability')}
      res <- clValid::clValid(obj = M.sub,
                            nClust = 3:8,
                            clMethods = 'kmeans',
                            validation = val,
                            maxitems = 10000,
                            nstart = 1000, iter.max = 30)
      stats <- data.table::as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')]
      colnames(stats) <- c('metric', 'k', 'clustering_method', 'score')
      stats$top_factors <- n
      return(stats)
    }))
    stats$cancer_type <- strsplit(experiment, "\\.")[[1]][1]
    stats$datatypes <- strsplit(experiment, "\\.")[[1]][2]
    return(stats)
  }))
  stats$method <- tool
  return(stats)
}))
parallel::stopCluster(cl)

@@

p1 <- ggboxplot(dt1, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
                     comparisons = comps) + 
  labs(title = 'Silhouette score', x = '') + 
  theme(legend.position = 'none')

dt2 <- cluster.stats[metric == 'APN'][datatypes == multi_omics_comb,
                                             .SD[which.min(score)],
                                             by = c('cancer_type', 'method', 'k')]
dt2$method <- toupper(gsub("_factors", "", dt2$method))
comps <- lapply(setdiff(unique(dt2$method), 'MAUI'), function(x) c('MAUI', x))
p2 <- ggboxplot(dt2, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE,  
                     comparisons = comps) + 
  labs(title = 'Cluster Stability Score (APN)', x = '') + 
  theme(legend.position = 'none')
